---
title: "RNA-seq exploratory quality control"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
params:
  data_root: "../results/rnaseq/exploratory_qc"
---

```{r}
root <- params$data_root
read_qc_table <- function(name) read.delim(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE,
  na.strings = character()
)
metrics <- read_qc_table(file.path("tables", "sample_metrics.tsv"))
filters <- read_qc_table(file.path("tables", "gene_filter_summary.tsv"))
scores <- read_qc_table(file.path("tables", "pca_scores.tsv"))
flags <- read_qc_table(file.path("tables", "advisory_flags.tsv"))
contrasts <- read_qc_table("contrasts.tsv")
groups <- unique(metrics$analysis_group)
```

## Methods

Gene-level Salmon estimated counts were filtered independently within each analysis group. Genes required a count of at least 10 in at least the smallest replicate-group size. Counts were rounded only for DESeq2 input; size factors and a blind variance-stabilizing transformation were used for unsupervised diagnostics. No sample was removed and no differential-expression test was fitted.

## Dataset summary

```{r}
knitr::kable(as.data.frame.matrix(table(metrics$species, metrics$bioproject)),
             caption = "Samples by species and BioProject")
knitr::kable(filters, caption = "Gene filtering by analysis group")
```

## Salmon QC

```{r}
salmon_columns <- intersect(c("sample_id", "species", "percent_mapped",
                              "mapping_flag", "library_types"), names(metrics))
knitr::kable(metrics[, salmon_columns, drop = FALSE],
             caption = "Salmon mapping and library diagnostics")
```

## Analysis groups

```{r, results='asis'}
for (group in groups) {
  cat("\n### ", group, "\n\n", sep = "")
  stems <- c("library_size", "detected_genes", "vst_distribution", "pca",
             "correlation_heatmap", "distance_heatmap")
  paths <- file.path(root, "figures", paste0(group, "_", stems, ".png"))
  knitr::include_graphics(paths)
  cat("\n")
}
```

## Advisory flags

```{r}
if (nrow(flags)) knitr::kable(flags) else cat("No advisory flags were emitted.")
```

## Joint BioProject diagnostics

These plots visualize BioProject separation only; they do not define cross-project biological contrasts.

```{r}
joint_plots <- file.path(
  root, "figures", c("wheat_joint_bioproject_pca.png",
                     "soybean_joint_bioproject_pca.png")
)
joint_plots <- joint_plots[file.exists(joint_plots)]
if (length(joint_plots)) knitr::include_graphics(joint_plots)
```

## Estimable contrasts

The following contrasts are documented for a later differential-expression workflow and were not fitted here.

```{r}
knitr::kable(contrasts)
```

## Reproducibility

```{r, results='asis'}
cat("```text\n")
cat(readLines(file.path(root, "session_info.txt")), sep = "\n")
cat("\n```\n")
```
